---
title: "Causal Inference"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
#####################################################################
library(ggdag)
coordLV <- list(
  x = c(sex = 0, age = 0, smoke = 1, height = 2, fev = 3),
  y = c(sex = 1, age = 2, smoke = 0, height = 2, fev = 0))
dagLV <- dagify(height ~ sex + age + smoke,
                smoke ~ sex + age,
                fev ~ height + sex + age + smoke,
                coords = coordLV,
                exposure = "smoke",
                outcome = "fev")
p1 <- ggdag(dagLV, text_col = "blue", node = FALSE, text_size = 5) + theme_dag_blank()
#####################################################################

library(BiocManager)
options(repos = BiocManager::repositories())

library(learnr)
library(tidyverse)
library(mosaic)
library(ggm)
library(MatchIt)
#####################################################################

FEV <- read.table('http://jse.amstat.org/datasets/fev.dat.txt')
colnames(FEV) <- c("age", "fev", "height", "sex", "smoke")
FEV <- as_tibble(FEV)
FEV <- FEV %>%
  mutate(sex = ifelse(!sex, "female", "male")) %>%
  mutate(smoke = ifelse(!smoke, "no", "yes"))

#####################################################################
# Exact matching
match.exact <- matchit(smoke ~ age + sex, data = FEV, 
                       estimand = "ATT",
                       method = "exact")
# Construct matched data
FEV.matched <- match.data(match.exact)
#####################################################################

tutorial_options(exercise.timelimit = 60)
```

## Preliminaries

This introductory tutorial is based on [Causal Inference in Introductory Statistics Courses](https://github.com/kfcaby/causalLab), see Cummiskey, K., Adams, B,. Pleuss, J.,  Turner, D., Clark, N. \& Watts, K. (2020). [*Causal Inference in Introductory Statistics Courses*, Journal of Statistics Education, 20(1)](https://doi.org/10.1080/10691898.2020.1713936).

Data is taken from Kahn, M. (2005). [*An exhalent problem for teaching statistics*. Journal of Statistics Education, 13(2)](https://doi.org/10.1080/10691898.2005.11910559).

<br>

*Note*: The R packages [mosaic](https://projectmosaic.github.io/mosaic/) with [ggformula](https://projectmosaic.github.io/ggformula/) are used for data analysis.

<br>

Please report any issues [here](https://github.com/luebby/ci-learnr/issues).

## Youth smoking

Nowadays we are sure that smoking is harmful. But this was not always clear and a naive data analysis may come to the opposite conclusion. 


In a series of papers (e.g. [Kahn, 2005](https://doi.org/10.1080/10691898.2005.11910559)) the question was analyzed if childhood smoking has an effect on  *forced expiratory volume*, a common measure of lung function.

###


```{r ethik, echo=FALSE}
question("Is it ethical in this case to conduct a randomized controlled trial?",
  answer("Yes.", message = "Negative consequences of smoking can be expected."),
  answer("No.", correct = TRUE, message = "Negative consequences of smoking can be expected.")
)
```


###

The data set `FEV` has the following structure:

```{r str}
str(FEV)
```

with variables:

- `age`: the age of the subject in years
- `fev`: forced expiratory volume (l)
- `height`: the height of the subject in inches
- `sex`: biological sex of the subject
- `smoke`: whether the subject had ever smoked or no

```{r head}
head(FEV)
```


```{r skalennvieau, echo=FALSE}
question("What is the measurement scale of `fev`?",
  answer("Nominal", message = "Values can be distinguished, ordered and as well differences as ratios are meaningful."),
  answer("Ordinal", message = "Values can be distinguished, ordered and as well differences as ratios are meaningful."), 
    answer("Interval", message = "Values can be distinguished, ordered and as well differences as ratios are meaningful."), 
    answer("Ratio", correct=TRUE, message = "Values can be distinguished, ordered and as well differences as ratios are meaningful.")
)
```

###

The following causal structure is assumed according to scientific background knowledge:

```{r p1, echo=FALSE}
p1
```

*Tip*: [DAGitty](http://dagitty.net/) and [ggdag](https://ggdag.netlify.app/) are nice tools for drawing and analyzing directed acyclic graphs.


## Smoking and forced expiratory volume

A boxplot of forced expiratory volume (`fev`) given smoking (`smoke`) shows the following result: 

```{r b1}
gf_boxplot(fev ~ smoke, data = FEV)
```

###

```{r median, echo=FALSE}
question("Which group has the higher median of `fev`?",
  answer("Non-smoker", message = "The median fev of non-smokers is $\\approx 2.5$ l, whereas for smokers it is $\\approx 3.2$  l."),
  answer("Smoker.", correct = TRUE, message = "The median fev of non-smokers is $\\approx 2.5$ l, whereas for smokers it is $\\approx 3.2$ l.")
)
```

###

Smoking as well as forced expiratory volume depend on sex.

Expand the code so that you can also condition on `sex`.

```{r bedingt, exercise = TRUE}
gf_boxplot(fev ~ smoke, data = FEV)
```

```{r bedingt-solution}
gf_boxplot(fev ~ smoke | sex, data = FEV)
```

###

The overall scene has not changed, smokers seem to have a higher forced expiratory volume.

###

Why?

## Age

Thinking about the data: height is associated with age -- especially while growing up:

```{r }
gf_point(height ~ age, data = FEV) %>%
  gf_smooth()
```

###

Expand the code to color the points by `smoke`:

```{r color, exercise = TRUE}
gf_point(height ~ age, data = FEV) %>%
  gf_smooth()
```

```{r color-solution}
gf_point(height ~ age, color = ~ smoke, data = FEV) %>%
  gf_smooth()
```

###

In childhood smoking: those who smoke are older than those who don't -- and also taller.

###

And of course there is an association between height and forced expiratory volume.

```{r}
gf_point(fev ~ height, data = FEV) %>%
  gf_smooth()
```


## Causal inference 

The assumed causal model of the data generating process is:

```{r p12, echo = FALSE}
p1
```


A naive modeling of forced expiratory volume by smoking gives the following result:

```{r}
lm(fev ~ smoke, data = FEV) %>%
  summary()
```

### 

Smoking seems to increase the forced expiratory volume ...

Why?

###

Because the non-causal *backdoor* path are open:

$$\text{smoke} \leftarrow \text{sex} \rightarrow \text{fev}$$

and

$$\text{smoke} \leftarrow \text{age} \rightarrow \text{fev}$$

Here age is in the middle of a *fork*.

These paths should be blocked by means of e.g. adjusting.

```{r chain, echo=FALSE}
question("Should we also adjust for `height`?",
  answer("Yes.", message = "`height` is on the causal path between `smoke` and `fev` (Mediator in a *chain*)."),
  answer("No.", correct = TRUE, message = "`height` is on the causal path between `smoke` and `fev` (Mediator in a *chain*).")
)
```

###

Give a formula which would allow to estimate the total causal effect of smoking on forced expiratory volume under the assumption that a linear model applies: 

```{r model, exercise = TRUE, eval=FALSE}
lm(fev ~ smoke + ___ + ___, data = FEV) %>%
  summary()
```

```{r model-solution}
lm(fev ~ smoke + sex + age, data = FEV) %>%
  summary()
```

###

```{r effekt, echo=FALSE}
question("Has smoking a negative total causal effect on the forced expiratory volume?",
  answer("Yes.", correct = TRUE, message = "The estimated coefficient is negative: $-0.153974$ (`Estimate` of `smokeyes`)."),
  answer("No.", message = "The estimated coefficient is negative: $-0.153974$ (`Estimate` of `smokeyes`).")
)
```

###

```{r pvalue, echo=FALSE}
question("Is the estimate ($\\hat{\\beta}_{\\text{smokeyes}}=-0.153974$) compatible to the model of no effect ($H_0: {\\beta}_{\\text{smokeyes}}=0$)?",
  answer("Yes.", message = "The probability in a model with ${\\beta}_{\\text{smokeyes}}=0$ to get a sample with an estimate as least as large as $\\hat{\\beta}_{\\text{smokeyes}}=|-0.153974|$ is $0.0487$ (p-value `Pr(>|t|)`."),
  answer("No.", correct = TRUE, message = "The probability in a model with ${\\beta}_{\\text{smokeyes}}=0$ to get a sample with an estimate as least as large as $\\hat{\\beta}_{\\text{smokeyes}}=|-0.153974|$ is $0.0487$ (p-value `Pr(>|t|)`.")
)
```

*Note*: The assumption of a linear model is due to the non-linear association between age and height as well as height and forced expiratory volume a questionable approximation.


### Simulation based inference

That the observed effect in the data is uncommon if the forced expiratory volume is independent of smoking can also be seen by a permutation test. For this we can use simulation based inference, i.e. simulating the distribution of $\hat{\beta}_{\text{smokeyes}}$ while shuffling smoke:

```{r permute, warning=FALSE , exercise = TRUE}
# Specify random number seed
set.seed(1896)

# Simulate permutation distribution
Nulldist <- do(1000) * lm(fev ~ shuffle(smoke) + sex + age, data = FEV)

# Visualize and compare with observed effect
gf_histogram( ~ smokeyes, data = Nulldist, center = 0, bins = 21) %>%
  gf_vline(xintercept = -0.153974)
```


## Summary

As Kari Lock Morgan puts it [here](https://askgoodquestions.blog/2020/07/27/56-questioning-causal-evidence/):

> Two key questions [we] should ask and answer when evaluating causal evidence:

>  - Do we have convincing evidence against "just random chance"?  Why or why not?

>  - Do we have convincing evidence against the groups differing to being with?  Why or why not?

For the data in this tutorial the smoking and non-smoking group differ by e.g. age. Therefore the naive comparison (without adjusting) was flawed.

The low p-value in the hypothesis test of ${\beta}_{\text{smokeyes}}=0$ gives us (weak) evidence against the difference just resulting from random chance. 




<br>

**Citation**: Karsten Lübke, Matthias Gehrke, Jörg Horst \& Gero Szepannek (2020) Why We Should Teach Causal Inference: Examples in Linear Regression With Simulated Data, Journal of Statistics Education, 28:2, 133-139, [https://doi.org/10.1080/10691898.2020.1752859](https://doi.org/10.1080/10691898.2020.1752859).

*Source*: [https://github.com/luebby/ci-learnr/](https://github.com/luebby/ci-learnr/)


## Appendix: Background

### Levels of causal inference

[Pearl (2019)](https://doi.org/10.1145/3241036) establishes a three-level hierarchy:

- **Association**: $P(y|x)$: Seeing: *what is?*, i.e., the probability of $Y=y$ given that we observe $X=x$.

- **Intervention**: $P(y|do(x))$: Manipulation: *what if?*, i.e., the probability of $Y=y$ given that we intervene and set the value of $X$ to $x$.

- **Counterfactuals**: $P(y_x|x',y')$: Imagining: *what if I had acted differently?*, i.e., the probability of $Y=y$ if $X$ had been $x$ given that we actually observed $x',y'$.

<br>

Other approaches to causal inference are e.g. within potential outcome framework, instrumental variables, regression discontinuity designs, Granger, natural experiments, ...

Compare also Cornfield's conditions and Bradford Hill criteria.

### Basic Elements of a DAG

- $X \rightarrow Y: \quad Y=f(X, U_Y)$ with some function $f(\cdot)$ and some exogenous $U$. 

- The value of $Y$ depends on $X$ - but the value of $X$ **not** on $Y$. 

- Causally there is no inverse function $f^{-1}(\cdot)$. My weight growths with my height but unfortunately my height not with my weight.

<br>



| Path                       | $X \rightarrow C \rightarrow Y$ | $X \leftarrow C \rightarrow Y$ | $X \rightarrow C \leftarrow Y$ 
| ---------------------------|---------------------------------|--------------------------------|------------------------------|
| Name                       | Chain                           | Fork                           | Collider         
| Association $X$ to $Y$     | Causal                          | Non-causal                     | None                       
| Role of $C$                | Mediator                        | Cause                          | Effect
| Adjusting $C$              | Blocks causal path              | Blocks non-causal path         | Opens biasing path


<br>

**Idea**: To estimate the change in $y$ if $x$ is changed: Block non-causal paths, open causal paths and don't open a biasing path.

*Tip*: [DAGitty](http://dagitty.net/) is a nice browser-based environment for creating, editing, and analyzing causal diagrams (directed acyclic graphs). See [Ankan et al. (2021)](https://doi.org/10.1002/cpz1.45).

## Appendix: Testable implications

The assumed causal structure of the data generating process implies some (conditional) independencies:

```{r p1indep, echo=FALSE}
p1
```

Here:

- Age independent of sex: $\texttt{age}{\perp\!\!\!\perp}\texttt{sex}$
- fev independent of age, conditional on height, sex and smoke: $\texttt{fev}{\perp\!\!\!\perp}\texttt{age}|\texttt{height}, \texttt{sex}, \texttt{smoke}$
- height independent of smoke, given age and sex: $\texttt{height}{\perp\!\!\!\perp}\texttt{smoke}|\texttt{age}, \texttt{sex}$

### Partial correlation tests

[DAGitty](http://dagitty.net/) returns additional to minimal sufficient adjustment sets for a given graph the R code for testable implications with `ggm::pcor.test()`:

```{r pcor,  exercise = TRUE}
# Convert (binary) categorical variables into logical variables
FEV2 <- FEV %>%
  mutate(smoke = (smoke=="yes")) %>%
  mutate(sex = (sex == "female"))

# R Code snippet from http://dagitty.net/
testImplications <- function( covariance.matrix, sample.size ){
  library(ggm)
  tst <- function(i){ pcor.test( pcor(i,covariance.matrix), length(i)-2, sample.size )$pvalue }
  tos <- function(i){ paste(i,collapse=" ") }
  implications <- list(c("age","sex"),
                       c("age","fev","height","smoke","sex"),
                       c("height","smoke","age","sex"))
  data.frame( implication=unlist(lapply(implications,tos)),
              pvalue=unlist( lapply( implications, tst ) ) )
}

testImplications( covariance.matrix = cov(FEV2), 
                 sample.size= nrow(FEV2) )
```
###

So there is a statistical discernible deviation of the partial correlation coefficient from $H_0: \texttt{fev}{\perp\!\!\!\perp}\texttt{age}|\texttt{height}, \texttt{sex}, \texttt{smoke}$ and the model should be refined. For $\texttt{age}{\perp\!\!\!\perp}\texttt{sex}$ and $\texttt{height}{\perp\!\!\!\perp}\texttt{smoke}|\texttt{age}, \texttt{sex}$ there is no statistical discernible deviation in the observed data.


## Appendix: Matching

In the matching approach, observations of smokers and nonsmokers are matched by their `age` and `sex` combination. Weights are assigned so that distribution of age and sex of nonsmokers (*Control*) match those of the smokers (*Treated*).

This can be done by the package [MatchIt](https://cran.r-project.org/package=MatchIt), see [Ho et al. (2011)](https://www.jstatsoft.org/article/view/v042i08).

```{r match, exercise = TRUE}
library(MatchIt)

# Exact matching
match.exact <- matchit(smoke ~ age + sex, data = FEV, 
                       estimand = "ATT",
                       method = "exact")

# Matching result
match.exact
summary(match.exact)

# Construct matched data
FEV.matched <- match.data(match.exact)
```

##

After matching the *Average Treatment Effect on Treated (ATT)* can be estimated:

```{r matchreg}
lm(fev ~ smoke, data = FEV.matched, weights = weights)
```

Again one can see a negative effect of smoking on the forced expiratory volume.


## Appendix: Learning more

With this tutorial we try to motivate you to consider bias and confounding in analyzing (causal) effects. More advanced and non-parametric methods exist. See for a start e.g.: 

- Cinelli, C., Forney, A., Pearl, J. (2020). [*A Crash Course in Good and Bad Controls*](https://ftp.cs.ucla.edu/pub/stat_ser/r493.pdf)
- Dablander, F. (2020). [*An Introduction to Causal Inference*](https://doi.org/10.31234/osf.io/b3fkw)
- Pearl, J., Glymour, M., \& Jewell, N. P. (2016). [*Causal inference in statistics: A primer*](http://bayes.cs.ucla.edu/PRIMER/)
- Rohrer,  J. (2018). [*Thinking Clearly About Correlations and Causation: Graphical Causal Models for Observational Data*](https://doi.org/10.1177%2F2515245917745629)

